Effective Reionization Likelihood from Planck 2018 Data (?)

### 2020-12-24 paper/rei.tex
Added text after restructuring

\subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
\label{sec:KDE}
\wh{The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data.}  In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~\ref{eq:xetommu}. Since $\mathbf{m}$ could take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
% 1.06 cut w/ physicality prior
% 1.072 no cut
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient. Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
\end{equation}
we demonstrate a successful recovery of the posterior distributions from direct MCMC analyses (using the Planck likelihood and varying all cosmological parameters) with significantly less computational time. We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

\subsection{\wh{Cumulative Optical Depth}}

\wh{Although reionization PCs are mainly a tool for model testing, as they do not reconstruct the ionization history $x_e(z)$ itself, they do provide model-independent constraints on  the cumulative Thomson optical depth}
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
$C_\ell^{EE}$ -- the higher the redshift, the
larger the relative contribution at higher $\ell$.

Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
\wh{decide whether to introduce models one-by-one or together - we could postpone the step model discussion until later without interrupting the flow}
For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
 \begin{equation}
x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
 \label{eqn:tanh}
 \end{equation}
 where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
 excludes finite ionization above its transition redshift by definition.

The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed.  \wh{The one subtlety when placing constraints on the cumulative optical depth is that the analysis also therfore allows
unphysical ionization fractions where $0<x_e<x_e^{\rm max}$  is not satisfied.}
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a physicality prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
%
Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.  \wh{In other words, no physical model is excluded by these priors.}  
% I think we don't impose the prior at all
% for KDE 
%
%They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.

@@

\wh{Think about whether the refutation of Marius' prior and should go here.  If not then we probably want a forward pointing sentence here about impact of priors in practice.}

@@

$m_1$ & $-$0.116 & 0.030 & 1.000 & 0.657 & -0.208 & -0.094 & 0.067 \\
$m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & -0.273 & 0.142 \\
$m_3$ & $$0.078 & 0.098 & -0.208 & 0.048 & 1.000 & 0.064 & -0.018 \\
$m_4$ & -0.077 & 0.131 & -0.094 & -0.273 & 0.064 & 1.000 & -0.195 \\
$m_5$ & 0.066 & 0.145 & 0.067 & 0.142 & -0.018 & -0.195 & 1.000 \\

@@

We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area. [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.] 
In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to an even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$ (also shown in Table~\ref{tab:PC_stats}.

\begin{equation}
 {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
 This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement. The Gaussian likelihood is even faster than the kernel density estimate, and may be used to sample well-constrained models near the peak of the PC distributions, while becoming possibly less accurate near the tails of the distributions.

In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{simall\_EE} and \texttt{srollv2}. The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]? \wh{we might make this point in the context of Marius - that $\tau$ is mainly a linear function of $m_1$, $m_2$, and hence flat priors in $m's$ are locally flat in $\tau$}  [Might be nice to overlay the tanh trajectories, no 1 sigma, 2 sigma business].

\wh{I wonder if the order should be reversed and talk about the model testing and KDE first, then the model independent cumulative optical depth?  That way the model stuff could all be introduced here and simulaneously compared with the fiducal results?  The flow could be that we first show more high z optical depth is allowed in the two step model but not FlexKnot and then talk about it being a model independent PC result, concluding that we are more robust than FlexKnot?  Of course I'm assuming that the 2 step model shows this...}

@@

As described in ~\refsec{KDE}, we turn the PC constraints from the Planck 2018 into an effective likelihood code to constrain arbitrary models of ionization history between $6 \leq z \leq \zmax$. This code is downloadable on Github at \url{} [give URL]. The initial release is offered for the Planck 2018 $\textsc{plik\_lite\_ TTTTTE} + \textsc{lowl} + \textsc{srollv2}$ likelihood combination with $\zmax = 30$.

@@

\begin{figure}
\includegraphics[width=0.40\textwidth]{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
\caption{Posterior of total optical depth $\tau$: full likelihood (solid) vs effective likelihood (dotted) with f = 0.14 [remove more curves?]. The results agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
}
\label{fig:tanh}
\end{figure}

@@

The steplike model used as the standard approach in CAMB describes the hydrogen and singly ionized helium reionization as a tanh function:

 \begin{equation}
x_e(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} + x_e^{\rm rec},
 \label{eqn:tanh_1D}
 \end{equation}

 with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$. Here $x_e^{\rm rec}$ is the ionization history from recombination only; the helium fraction
 \beq
 f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
 \eeq
 is the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density. In addition, the doubly ionized helium reionization is parameterized by an additional tanh function in redshift centered at $z_{\rm He} = 3.5$ with width $\Delta z = 0.5$.

In \reffig{tanh_1D} we show the posterior distribution of the $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which all six $\Lambda$CDM parameters were varied including $\tau$. The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing. We expect the smoothing to be at roughly the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE.

\begin{itemize}
    \item (here?) Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.
    \item{Might also want to address this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true? Then transition to the following point}
\end{itemize}

 %We take here $z_{\rm re}= 9.85$, corresponding the chain maximum likelihood (ML) model ($\tau = 0.0765$) from \S \ref{sec:MCMC},  for illustrative purposes.   Projected onto 5 PCs and resummed into $x_e(z)$, Eq.~(\ref{eq:mmutoxe}) yields a poor reconstruction of the ionization history itself.
 %Nonetheless as we shall see in Fig.~\ref{fig:clee}, the PC decomposition provides an excellent representation of the polarization power spectrum.

@@

The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
 \bea
x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
 \label{eqn:tanh_highz}
 \eea
where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$ as in the first example, but now we choose $\Delta z_1 = 0.015\, (1+z_{\rm re})$ instead of the usual $\Delta z_1 = 0.5$, to provide sharper distinctions between the two steps.
We choose the second step at $z_{\rm t}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
We recover the standard tanh model in the limit when $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$.
Therefore this model adds a single parameter $\xemin$ to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$. We show an example
of this two-step model in Fig.~\ref{fig:two_step_model}.

@@

To obtain posterior distributions with flat $\tau$ priors, we convert the model to be parameterized by low and high-redshift optical depth, $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$, 
each defined as in Eq.~\refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ respectively. We choose the split to be $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data. 
We sample in the $(\taulo, \tauhi)$ parameter space and the corresponding ($\zre, \xemin$) is found through an iterative search similar to that used for finding $\zre$ given total $\tau$ in the canonical tanh model. 
The priors are chosen to be $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where $\tau_{\rm max} = 0.35$, and $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth for both the effective and full likelihood runs, 
except that we have chosen $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. 
Note that $\taulo \lesssim 0.038$ roughly corresponds to $\zre < 6$ in the canonical tanh model. For the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$. 
So to make the comparison apples-to-apples, we further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the effective reionization likelihood chains and the full likelihood chains.

In ~\reffig{two_parameter_model_2D}, we compare the posteriors of $\taulo$ and $\tauhi$ obtained with these two likelihood codes, which show good agreement. Again we used $f = 0.14$ here for the KDE computation. [add timing results here?] Note that the truncation of the 2D constraint ellipses on the lower end of $\taulo$ is curved cut rather than a straight line, because $\zre$ cut we performed corresponds to a different value of $\taulo$ for a different $\tauhi$.

The marginalized 1D constraints are $\taulo = XX \pm XX$, $\tauhi = XX \pm XX$ for both runs. The 95\% C.L. upper limit on $\tauhi$ is $\tauhi < XX$ is consistent with the $\tau(15, 30) < XX$ results we got from the PC constraints. [is it? double check] This model of course do not capture the larger space of physical models with high-$z$ ionization like the FlexKnot or PC methods would, but consists of a subset of models for which the high-redshift optical depth results must still be consistent with the constraints from these more general methods.

To test this, we take the best-fit two-parameter model $(\taulo, \tauhi) = (XX, XX)$ and show its cumulative optical depth $\tau(z, \zmax)$ in ~\reffig{plot_taugtz_two_step_best_fit}. Compared to the PC constraints (shaded regions), this model is within the 95\% upper limit at all redshifts. According to the FlexKnots constraints in Ref.~\cite{}, however, this model would have been ruled out given that $\tau(15, 30) < 0.007$ (95\% C.L.) with the FlexKnots results. [provide comment ...; any further discussion on literature results with FlexKnots?]

[add a conclusive paragraph or sentence]

>>>>

\subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
\label{sec:KDE}
The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data. In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~\ref{eq:xetommu}. Since $\mathbf{m}$ could take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
% 1.06 cut w/ physicality prior
% 1.072 no cut
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient. Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
\end{equation}
we demonstrate a successful recovery of the posterior distributions from direct MCMC analyses (using the Planck likelihood and varying all cosmological parameters) with significantly less computational time. We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

An even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
%
\begin{equation}
 {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
%
We will compare its performance to the KDE likelihood for the Planck 2018 data in the two examples of section~\ref{sec:effective_likelihood} as well.

\subsection{Cumulative Optical Depth}

Although reionization PCs are mainly a tool for model testing, as they do not reconstruct the ionization history $x_e(z)$ itself, they do provide model-independent constraints on  the cumulative Thomson optical depth
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
$C_\ell^{EE}$ -- the higher the redshift, the
larger the relative contribution at higher $\ell$.

Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also therfore allows
unphysical ionization fractions where $0<x_e<x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a physicality prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
%
Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.  In other words, no physical model is excluded by these priors.
% I think we don't impose the prior at all
% for KDE 
%They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.

@@

\wh{want a forward pointing sentence here about impact of priors in practice.}

@@

$m_1$ & $-$0.116 & 0.030 & 1.000 & 0.657 & $-$0.208 & $-$0.094 & 0.067 \\
$m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & $-$0.273 & 0.142 \\
$m_3$ & $$0.078 & 0.098 & $-$0.208 & 0.048 & 1.000 & 0.064 & $-$0.018 \\
$m_4$ & $-$0.077 & 0.131 & $-$0.094 & $-$0.273 & 0.064 & 1.000 & $-$0.195 \\
$m_5$ & 0.066 & 0.145 & 0.067 & 0.142 & $-$0.018 & $-$0.195 & 1.000 \\

@@

We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area. [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.] 
In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to the simpler Gaussian likelihood of Eq.~\ref{eq:gaussian}. The exact values of the PC mean and $\bar{\bf m}$ and covariance $\bf C$ are shown in Table~\ref{tab:PC_stats}.

 This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement. The Gaussian likelihood is even faster than the kernel density estimate. As shall see in the next section for two explicit examples, the Gaussian likelihood shows generally good agreement with the full KDE result, but is unable to capture, as expected given its symmetric nature, any skewness in the PC distribution that may affect parameter constraints.

In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{simall\_EE} and \texttt{srollv2}. The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]? \wh{we might make this point in the context of Marius - that $\tau$ is mainly a linear function of $m_1$, $m_2$, and hence flat priors in $m's$ are locally flat in $\tau$}.

@@

As described in ~\refsec{KDE}, we turn the PC constraints from the Planck 2018 into an effective likelihood code to constrain arbitrary models of ionization history between $6 \leq z \leq \zmax$. This code is available on GitHub at \url{} [give URL]. The initial release is offered for the Planck 2018 $\textsc{plik\_lite\_TTTTTE} + \textsc{lowl} + \textsc{srollv2}$ likelihood combination with $\zmax = 30$.

@@

\begin{figure}
\includegraphics[width=0.40\textwidth]
%{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
{plots/plot_tanh_gaussian_MCMC.png}
\caption{Posterior of total optical depth $\tau$: full likelihood (solid) vs effective likelihood (dotted) with f = 0.14 [remove more curves?]. The results agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
}
\label{fig:tanh}
\end{figure}

@@

The steplike model used as the standard approach in CAMB describes the hydrogen and singly ionized helium reionization as a tanh function:
%
 \begin{equation}
x_e(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} + x_e^{\rm rec},
 \label{eqn:tanh_1D}
 \end{equation}
 %
 with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$. Instead of using $\Delta z = 0.5$ as in the standard setting by camb, we adopt a smaller width $\Delta z = 0.015(1+z)$. Here $x_e^{\rm rec}$ is the ionization history from recombination only; the helium fraction
 \beq
 f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
 \eeq
 is the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density. In addition, the doubly ionized helium reionization is parameterized by an additional tanh function in redshift centered at $z_{\rm He} = 3.5$ with width $\Delta z = 0.5$.

In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which all six $\Lambda$CDM parameters were varied including $\tau$. 
Because the PC method cannot accurately represent models that are not fully ionized below $z = 6$, we apply a prior $\zre < 6.1$ to both chains given the transition width of about $0.015(1+z) \approx 0.1$ at $z = 6$. The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing. We expect the smoothing to be roughly at the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE. Indeed, the KDE result $\tau = 0.0588 \pm 0.0066$ has a negligibly lower mean ($0.05\sigma$) and a 6.5\% larger standard deviation (as expected) than the full likelihood result $\tau = 0.0591 \pm 0.0062$;.

Instead of the effective likelihood, we can also use its Gaussian approximation shown in Eq.~\ref{eq:gaussian}. The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain. [add timing here]
%and amounts to one evaluation of the multi-variate Gaussian function per sample in the reionization model parameter space. 
We obtain for the tanh model $tau = 0.0594 \pm 0.0063$. While it is a good approximation (the mean and standard deviation are similar to those from the full likelihood and the effective likelihood using KDE), we see in Fig.~\ref{fig:tanh} that the peak of the distribution for the Gaussian likelihood is offset from the other two. This is because the Gaussian likelihood is by definition symmetric in the PC space and so it cannot capture any skewness in the PC distribution. As a result the tanh $\tau$ constraints are mostly symmetric too, $\tau = 0.0594_{-0.0064}^{+0.0064}$ if we consider the 68\% lower and upper limits, whereas the full KDE result $\tau = 0.0588_{-0.0072}^{+0.0060}$ is able to capture the skewness in the distribution $\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}.

\ch{[double check all these numbers again, write script to output in a table automatically]}.

[Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.]

[Note sure if it is related to this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true? Then transition to the following point]

We also show in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh} the cumulative optical depth obtained from directly integration the exact tanh model in the full likelihood chains vs that from the 5-PC representation of the tanh models in the effective likelihood chains. [describe more once plot is made]. From these constraints it is also clear that the tanh model does not allow for high redshift ionization by construction, while the PC chains sampling the entire space of physical models would still allow for some high-$z$ ionization in the Planck 2018 data, as we shall see later in in Section~\ref{sec:results}.

@@

The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
 \bea
x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
 \label{eqn:tanh_highz}
 \eea
where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$ where $\Delta z_1 = 0.015\, (1+z_{\rm re})$ just as in the first example for a sharper step.
We choose the second step at $z_{\rm t}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$. We would recover the standard tanh as $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$. We show an example
of this two-step model in Fig.~\ref{fig:two_step_model}.

@@

In practice, we sample the model in  $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$ space to obtain posterior distributions with flat $\tau$ priors. They are defined as in Eq.~\refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data. [double check this too.]
Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given total optical depth $\tau$. 
For both the full likelihood and effective likelihood runs, we adopt the priors $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where  $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth, and $\tau_{\rm max} = 0.35$. 
%except that $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. [not sure if this last sentence is needed; since we put a zre prior anyways.]
We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains. Here we used again $f = 0.14$ in the KDE computation.

In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$. They agree very well. Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$; for the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$. This is why the cut on the 2D constraint ellipses at low $\taulo$ is of a curved shape rather than a straight cut, because the $\zre$ prior corresponds to a different value of $\taulo$ for a different $\tauhi$.

[add timing results here?]

[comment on effective likelihood w/ KDE vs Gaussian likelihood?]

The marginalized 1D constraints from the effective likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$ are consistent with those of the full likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$. The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < XX$ for the effective likelihood and $\tau(15, 30) < XX$ for the full likelihood.
